BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-02-05 paper/paper.qmd
first pass of section 4.1

### Empirical evidence for

>>>>

### Empirical Evidence for Sub-Block Descent and Momentum

To showcase that the combination of these two tricks can speed up convergence, we will benchmark them by factorizing a random $10\times 10$ tensor (a matrix) with rank $3$. The Julia code is shown below, and the results are shown in @tbl-subblock-momentum-results.

```julia
using BenchmarkTools
using BlockTensorDecomposition

fact = BlockTensorDecomposition.factorize

options = (
    :rank => 3,
    :tolerence => (1, 0.03),
    :converged => (GradientNNCone, RelativeError),
    :δ => 0.9,
)

n_subblock_n_momentum(Y) = fact(Y;
    do_subblock_updates=false,
    momentum=false,
    options...
)

y_subblock_n_momentum(Y) = fact(Y;
    do_subblock_updates=true,
    momentum=false,
    options...
)

n_subblock_y_momentum(Y) = fact(Y;
    do_subblock_updates=false,
    momentum=true,
    options...
)

y_subblock_y_momentum(Y) = fact(Y;
    do_subblock_updates=true,
    momentum=true,
    options...
)

I, J = 10, 10
R = 3

@benchmark n_subblock_n_momentum(Y) setup=(Y=Tucker1((I, J), R))
@benchmark n_subblock_y_momentum(Y) setup=(Y=Tucker1((I, J), R))
@benchmark y_subblock_n_momentum(Y) setup=(Y=Tucker1((I, J), R))
@benchmark y_subblock_y_momentum(Y) setup=(Y=Tucker1((I, J), R))

performance_increase(old, new) = (old - new) / new * 100
```

The code `Tucker1((I, J), R)` produces a random $I\times J$ rank-$R$ matrix by generating two matrices $A\in\mathbb{R}^{I \times R}$ and $B\in\mathbb{R}^{R \times J}$ with standard normal entries, and multiplies them together.

|                   | **No Momentum**            | **Yes Momentum**           |
|-------------------|----------------------------|----------------------------|
| **No Sub-Block**  | 48.843 ms                  | 45.738 ms (6.7887% faster) |
| **Yes Sub-Block** | 27.473 ms (77.785% faster) | **24.350 ms (100.59% faster)** |

: Summary of median times to factorize a random $10\times 10$ rank-3 matrix under different methods. The performance increase is given by the formula $(\text{old} - \text{new}) / \text{new}$. {#tbl-subblock-momentum-results}

In @tbl-subblock-momentum-results, you can see that having both sub-block descent and momentum yields the fastest factorization. Moreover, the performance increase is *more* than simply the performance increases obtained by exclusively sub-block or momentum alone.^[The expected performance increase if sub-block descent and momentum where independence would be $(1+0.067887)(1+0.77785) = 1.89854$ or only 89.854% faster.] This suggests that there is synergy with these two methods and are best used together.
